Structural Insights 2023: A Linear Analysis of Apartment Age and Safety Scores in Toronto

### 2024-03-17 paper/paper.qmd
added posterior check

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
```

@@

Our study's estimand is the effect of building age on evaluation scores in Toronto. We investigate how the age of buildings influences their assigned evaluation scores, reflecting their compliance with safety standards and overall condition. Utilizing a dataset from the Toronto Open Data portal, which includes 1,758 records, we apply a simple linear regression model to forecast evaluation scores for Toronto residences in 2023. By conducting this analysis, we aim to quantify the impact of building age on evaluation scores and understand its implications for urban housing safety and maintenance.

@@

In this research, we examine apartment data sourced using the opendatatoronto package [@Toronto_2023]. The dataset encompasses 1,758 records (1,748 records after cleaning), focusing on specific variables: the ward of each apartment, its evaluation score, and the construction year. The data covers evaluations across Toronto's 25 wards, with scores varying between a minimum of 0 and some achieving the maximum score of 100. The study includes apartments that are over three stories in height and contain at least 10 units. These apartments span a wide range of construction years, from as early as 1805 to as recent as 2023.

@@

## Data Exploration:

@@

- **Count**: The dataset encompasses a total of 1,748 buildings, indicating a robust sample size for statistical analysis.

@@

The objective of our model is dual: to discern the relationship between building age and safety evaluation scores and to quantify this relationship's strength and direction. The study utilizes a simple linear regression model to analyze the dataset obtained from the Toronto Open Data portal.

@@

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{AGE}_i \\
\beta_0 &\sim \mbox{Normal}(85, 10) \\
\beta_1 &\sim \mbox{Normal}(0, 0.05)
\end{align}

The model is implemented in R using the rstanarm package, which allows us to incorporate prior beliefs about the parameters and to estimate the posterior distributions. For the intercept $\beta_0$ and slope $\beta_1$ , we choose normal prior distributions centered around our initial estimates based on exploratory data analysis. The `stan_glm` function from `rstanarm` is utilized to carry out the Bayesian regression, providing a framework for inference that incorporates both the data and our prior knowledge.

The priors are chosen to reflect reasonable beliefs about the parameters before observing the data. The intercept prior reflects an expectation that a new building will have a high safety score around 90, with a wide spread to accommodate uncertainty. Similarly, the slope prior indicates an expected small negative impact of age on the safety score, acknowledging some uncertainty in the precise effect. The `stan_glm` function combines these priors with the data to estimate the posterior distributions of the model parameters.

@@

**Intercept** ($\beta_0$):  
The model's intercept, $\beta_0$, has an estimate of 93.364. This represents the baseline evaluation score for a hypothetical newly constructed building. The intercept suggests that if a building were constructed in the current year, its expected evaluation score would be around 93.364, assuming all other factors remain constant.

**Slope** ($\beta_1$):  
The slope parameter, $\beta_1$, has an estimate of -0.097. This indicates that, on average, the evaluation score is expected to decrease by approximately 0.097 points for each additional year of a building's age. This negative value suggests a trend where older buildings tend to have lower evaluation scores compared to newer ones.

@@

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times x_i \\
\beta_0 &= 93.364 \\
\beta_1 &= -0.097
\end{align}

In this model, $y_i$ represents the predicted evaluation score for the $i$-th building, $x_i$ represents the age for the $i$-th building, $\mu_i$ is the mean of the normal distribution for the $i$-th observation, $\beta_0$ is the intercept, and $\beta_1$ is the slope representing the change in evaluation score per year of building age.

@@

**1. Minimum Residual**: The minimum residual is approximately -69.700, indicating that there is at least one building for which the model's predicted evaluation score was lower than the actual score by approximately 69.700 points. This could be an outlier or a building with exceptional characteristics not captured by the model.

**2. First Quartile Residual**: The first quartile (Q1) of the residuals is approximately -4.440, meaning that 25% of the buildings have evaluation scores that are less than 4.440 points below the model's prediction for their age.

**3. Median Residual**: The median residual is approximately 1.687, suggesting a slight positive bias in the model's predictions, as the median of the error distribution is above zero. This indicates that, on average, the model tends to slightly overestimate the evaluation scores.

**4. Third Quartile Residual**: The third quartile (Q3) is approximately 5.860, indicating that 75% of the buildings have evaluation scores within 5.860 points of the model's predictions or better. This shows that the model provides reasonably accurate predictions for the majority of buildings.

**5. Maximum Residual**: The maximum residual is approximately 15.715, suggesting that there is at least one building for which the model's predicted evaluation score was higher than the actual score by approximately 15.715 points. This could be another outlier or a building that performed exceptionally well.

@@

## Plot for the linear model
```{r fig-linear-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="plot for the linear relationship between building age and evaluation score"}
# Code created with the help from ChatGPT
# Enhanced scatter plot with regression line
improved_plot <- ggplot(cleaned_data, aes(x = AGE, y = CURRENT.BUILDING.EVAL.SCORE)) +
  geom_point(alpha = 0.7, size = 1) +  # Adjust transparency and size of points
  geom_smooth(method = "lm", color = "blue", size = 1) +  
  labs(
    x = "Building Age",
    y = "Building Evaluation Score",
    title = "Impact of Building Age on Evaluation Score"
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none"  # Hide the legend if not necessary
  )

@@

This study embarked on a journey to understand the relationship between the age of apartment buildings in Toronto and their safety evaluation scores. Leveraging a dataset of 3,435 records from the Toronto Open Data portal, we employed simple linear regression analysis to unearth a significant inverse correlation between building age and safety scores. Specifically, our findings indicate that older buildings in Toronto tend to receive lower safety evaluation scores. This observation underscores the potential risks associated with aging infrastructure, aligning with the broader literature on building safety and maintenance.

@@

# Appendix

@@

>>>>

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
library(rstanarm)
```

@@

Our study's estimand is the effect of building age on evaluation scores in Toronto. We investigate how the age of buildings influences their assigned evaluation scores, reflecting their compliance with safety standards and overall condition. Utilizing a dataset from the Toronto Open Data portal [@Toronto_2023], which includes 1,758 records, we apply a simple linear regression model to forecast evaluation scores for Toronto residences in 2023. By conducting this analysis, we aim to quantify the impact of building age on evaluation scores and understand its implications for urban housing safety and maintenance.

@@

In this research, we examine apartment data sourced using the opendatatoronto package [@Toronto_2023]. The dataset encompasses 1,758 records (1,748 records after cleaning), focusing on specific variables: the ward of each apartment, its evaluation score, and the construction year. The data covers evaluations across Toronto's 25 wards, with scores varying between a minimum of 0 and some achieving the maximum score of 100. The study includes apartments that are over three stories in height and contain at least 10 units. These apartments span a wide range of construction years, from as early as 1805 to as recent as 2023. For description of the data cleaning process and data validation checks performed, please refer to @sec-appendix (Appendix).

@@

## Data Exploration: {#sec-exploration}

@@

- **Count**: The analysis dataset encompasses a total of 1,748 buildings, indicating a robust sample size for statistical analysis.

@@

The objective of our model is dual: to discern the relationship between building age and safety evaluation scores and to quantify this relationship's strength and direction. The study utilizes a simple linear regression model to analyze the dataset obtained from the Toronto Open Data portal. Additional details and diagnostics for this model are included in @sec-posterior.

@@

\begin{align}
y_i | \mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times x_i \\
\beta_0 &\sim \mbox{Normal}(90, 10) \\
\beta_1 &\sim \mbox{Normal}(-0.1, 0.05) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

The model is implemented in R using the `rstanarm` package, which allows us to incorporate prior beliefs about the parameters and to estimate the posterior distributions. For the intercept $\beta_0$ and slope $\beta_1$, we choose normal prior distributions centered around our initial estimates based on exploratory data analysis (@sec-exploration). The `stan_glm` function from `rstanarm` is utilized to carry out the Bayesian regression, providing a framework for inference that incorporates both the data and our prior knowledge.

The priors are chosen to reflect reasonable beliefs about the parameters before observing the data. The intercept prior reflects an expectation that a new building will have a high safety score around 90, with a wide spread to accommodate uncertainty. Similarly, the slope prior indicates an expected small negative impact of age on the safety score, acknowledging some uncertainty in the precise effect. The standard deviation $\sigma$ is given an exponential prior with a rate of 1, reflecting a belief in the likelihood of smaller variability in evaluation scores. The `stan_glm` function combines these priors with the data to estimate the posterior distributions of the model parameters.

@@

**Intercept** ($\beta_0$):  
The model's intercept, $\beta_0$, has an estimate of 93.607. This represents the baseline evaluation score for a hypothetical newly constructed building. The intercept suggests that if a building were constructed in the year for evaluation, its expected evaluation score would be around 93.607, assuming all other factors remain constant.

**Slope** ($\beta_1$):  
The slope parameter, $\beta_1$, has an estimate of -0.100. This indicates that, on average, the evaluation score is expected to decrease by approximately 0.100 points for each additional year of a building's age. This negative value suggests a trend where older buildings tend to have lower evaluation scores compared to newer ones.

@@

\begin{align}
y_i &= 93.607 - 0.100 x_i + \varepsilon_i
\end{align}

The above model equation serves as a predictive tool, estimating a building's evaluation score based on its age. In this equation, $y_i$ is the estimated evaluation score, $x_i$ represents the age of the building, $\varepsilon_i$ captures the residual effect—factors influencing the score that are not explained by the building's age. This implies that even among buildings of similar ages, scores can differ based on variables like maintenance quality, renovations, or compliance with updated building codes. The equation recognizes that while age is a significant predictor, the complexity of real-world conditions means there's always an element of variability unexplained by age alone.

@@

**1. Minimum Residual**: The minimum residual is approximately -69.676, indicating that there is at least one building for which the model's predicted evaluation score was lower than the actual score by approximately 69.676 points. This could be an outlier or a building with exceptional characteristics not captured by the model.

**2. First Quartile Residual**: The first quartile (Q1) of the residuals is approximately -4.377, meaning that 25% of the buildings have evaluation scores that are less than 4.377 points below the model's prediction for their age.

**3. Median Residual**: The median residual is approximately 1.705, suggesting a slight positive bias in the model's predictions, as the median of the error distribution is above zero. This indicates that, on average, the model tends to slightly overestimate the evaluation scores.

**4. Third Quartile Residual**: The third quartile (Q3) is approximately 5.821, indicating that 75% of the buildings have evaluation scores within 5.821 points of the model's predictions or better. This shows that the model provides reasonably accurate predictions for the majority of buildings.

**5. Maximum Residual**: The maximum residual is approximately 15.835, suggesting that there is at least one building for which the model's predicted evaluation score was higher than the actual score by approximately 15.835 points. This could be another outlier or a building that performed exceptionally well.

@@

## Plot for the linear model
```{r fig-linear-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="plot for the linear relationship between building age and evaluation score"}
# Code created with the help from ChatGPT
# Enhanced scatter plot with regression line
improved_plot <- ggplot(cleaned_data, aes(x = AGE, y = CURRENT.BUILDING.EVAL.SCORE)) +
  geom_point(alpha = 0.7, size = 1) +  # Adjust transparency and size of points
  geom_smooth(method = "lm", color = "blue", size = 1) +  
  labs(
    x = "Building Age",
    y = "Building Evaluation Score",
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.position = "none"  # Hide the legend if not necessary
  )

@@

This study embarked on a journey to understand the relationship between the age of apartment buildings in Toronto and their safety evaluation scores. Leveraging a dataset of 1,758 records from the Toronto Open Data portal, we employed simple linear regression analysis to unearth a significant inverse correlation between building age and safety scores. Specifically, our findings indicate that older buildings in Toronto tend to receive lower safety evaluation scores. This observation underscores the potential risks associated with aging infrastructure, aligning with the broader literature on building safety and maintenance.

@@

# Appendix {#sec-appendix}

@@

## Posterior Predictive Check {#sec-posterior}

```{r fig-posterior-plot, echo=FALSE, fig.cap="comparison between observed evaluation scores with those generated by our Bayesian model"}
# This code generates plot that compare the distribution of the observed data to the distributions implied by the posterior predictive distribution of the model
pp_check(bayesian_model) + 
  theme_classic() +
  theme(legend.position = "bottom")
```
From @fig-posterior-plot, we observe that the simulated data from the posterior distribution (represented by the lighter lines) closely match the distribution of the observed data (depicted by the solid line). This suggests that our model has a good fit to the observed data and that the chosen priors and model structure can reasonably replicate the variations in evaluation scores across buildings. The spread of the lighter lines around the solid line also highlights the degree of uncertainty in our predictions, which is a natural aspect of modeling real-world phenomena.

```{r fig-comparison-plot, echo=FALSE, message=FALSE, fig.cap="Comparison of posterior and prior distributions for model parameters"}
posterior_vs_prior(bayesian_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@fig-comparison-plot displays the median and spread of prior and posterior distributions for our model's parameters. It shows the posterior for the slope, related to the building's age, is quite similar to the prior, meaning our initial guess was on target. For the intercept and sigma, the data has slightly adjusted our estimates, reflecting its informative value. Overall, the data confirmed our prior thoughts about the effect of age on the building's evaluation score without much alteration.

\newpage
